Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-27 paper.tex
Updates from Overleaf

\includegraphics[width=0.4\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
\\ % &
% \multicolumn{2}{c}{
\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
% }
\end{tabularx}
\caption{\label{fig:pango-single-origin-graph}
Single origin Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show where the ARG continues. Nodes are coloured by Nextclade Pango designation; smaller symbols are nodes inserted by  sc2ts (i.e. not samples); their Pango designation has been imputed where possible. (B) Details for the single-origin Pango X lineages in the long ARG. (C) XAG origins in the long ARG.}
\end{figure}

In the long ARG, the sequences of 16 Pango recombinant lineages trace back to a single recombination node, as summarised in Table~\ref{tab:pango-single-origin}.
[FIXME UPDATE THESE]
We can see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the samples given that Pango designation descendand from a single recombination node.

@@

[Discuss panel C where XAG shares a common recombinant origin with samples from several other X lineages]
All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade. Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the sample nodes, which can be converted to EPI\_ISL indentifiers using Supplemeentary table XXX )

@@

Example lineage subgraph
% \includegraphics[width=0.4\textwidth]{figures/Pango_XD_nxcld_tight_graph.pdf}

@@

\begin{tabular}{llll}
\toprule
Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
\midrule
BA.1.15 & BA.2  &9867  &10198\\
BA.1.15 & BA.2  &9867  &10198\\
BA.1.1  &BA.2  &10451  &11537\\
BA.1.1.5 & BA.2 &10448 & 11537\\
BA.1.15 & BA.2  &10448  &12880\\
BA.1.17 & BA.2  &10448 & 11537\\
BA.1  &BA.2.9  &10448 & 11537\\
BA.1  &BA.2  &12764  &13195\\
\bottomrule
\end{tabular}
% BA.3  &BA.1.17  &22674  &22674\\
% BA.1.1.15 Recombinant 10448 11537
% BA.1.1 Recombinant 10030 10447
% Recombinant BA.2 17411 19955
\end{tabular}
\caption{A. Example subgraph for XD in the long ARG, which shows . B. details of inferred breakpoints for different recombination events inferred for XE samples (170 samples and 12 inferred recombination nodes; some are very similar (Official: BA.1* \& BA.2, 10448–11287)}
\end{figure}

In Figure XXX we give the details of these multiple pango lineages, along with an example lineage subgraph. As with the subgraphs in Figure~\ref{tab:pango-single-origin}, the equivalent XD plot with GISAID Pango designations, along with identifying labels, is shown for comparison in Supplementary Figure XXX

@@

As well as cases where a single recombination event explains several X lineages (such as XAG above), many of the Pango X lineages involve more complex origins, in which samples which have not been previously identified as recombinants also descend from originating recombination events. In addition these complex topologies often include extra recombination nodes within the topology that descends from the main originating event. An example is XB in the long ARG, plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC). However, in this subgraph, there is only a single sample, identified as XB by Nextclade, which is descended from these
nested recombination events; moreover, this sample is not identified as XB by GISAID (Supplementary Figure~\ref{}). We suggest that investigating cases of complex origins, and identifing which (if any) of the nested, further recombination events may be artifactual, is an important area of future research.

>>>>

\includegraphics[width=0.6\textwidth]{figures/Pango_XA_nxcld_tight_graph.pdf}
\\ % &
% \multicolumn{2}{c}{
\includegraphics[width=\textwidth]{figures/Pango_XAG_nxcld_tight_graph.pdf}
% }
\end{tabularx}
\caption{\label{fig:pango-single-origin-graph}
Single origin Nextclade Pango X lineages. (A) Subgraph for XA in the long ARG: all XA samples and their ancestral lineages are shown outwards to the nearest sampled viral genome; dotted lines show where the ARG continues. Nodes are coloured by Nextclade Pango designation; smaller symbols are nodes inserted by  sc2ts (i.e. not samples); their Pango designation has been imputed where possible. (B) Details for the single-origin Pango X lineages in the long ARG. (C) XAG origins in the long ARG.}
\end{figure}

In Figure~\ref{fig:pango-single-origin-graph}A we illustrate the case of XA. We construct a small subgraph of the long ARG, tracing routes from Nextclade-identified XA samples to the closest other sample strains in our dataset. Where a route passes upwards through a recombination node, we only show the continuing parental lineages to the most recent sample ancestors, rather than descending the ARG downward to cousins of the recombinant parents. Dotted lines show where this subgraph links to other nodes within the ARG. It is clear from the subgraph that all the XA samples trace to a single originating recombination node.

Table~\ref{tab:pango-single-origin} summarises the sequences of all 16 of the Pango recombinant lineages that trace back to a single recombination node in the long ARG.
[FIXME UPDATE THESE]
We can see that XA, XAC, XAE, and XF form monophyletic clades, in which all of the samples given that Pango designation descendand from a single recombination node.

@@

[Discuss panel C where XAG shares a common recombinant origin with samples from several other X lineages]
All XAG samples share a single origin; however this is also shared with all XAA samples, and includes some, but not all, of the samples identified as XAB, XQ, and XU by Nextclade. Note that if the GISAID designations are used, many of the samples marked here as XAB are reclassified as BA.2, and XAG becomes fully monophyletic (supplementary figure \ref{fig:pango_XAG_gisaid_graph}; this also gives identifying labels to the sample nodes, which can be converted to EPI\_ISL indentifiers using Supplementary table XXX )

@@

\includegraphics[width=0.45\textwidth]{figures/Pango_XD_nxcld_tight_graph.pdf}

@@

\begin{tabular}{llll}
\toprule
Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
\midrule
BA.1.15 & BA.2  &9867  &10198\\
BA.1.15 & BA.2  &9867  &10198\\
BA.1.1  &BA.2  &10451  &11537\\
BA.1.1.5 & BA.2 &10448 & 11537\\
BA.1.15 & BA.2  &10448  &12880\\
BA.1.17 & BA.2  &10448 & 11537\\
BA.1  &BA.2.9  &10448 & 11537\\
BA.1  &BA.2  &12764  &13195\\
\bottomrule
\end{tabular}
% BA.3  &BA.1.17  &22674  &22674\\
% BA.1.1.15 Recombinant 10448 11537
% BA.1.1 Recombinant 10030 10447
% Recombinant BA.2 17411 19955
\end{tabular}
\caption{\label{fig:multiple_origins_graph} A. Example subgraph for XD in the long ARG, which shows two origins for the four Nextclade-labelled XD samples in the long ARG. B. details of inferred breakpoints for different recombination events inferred for XE samples (170 samples and 12 inferred recombination nodes; some are very similar (Official: BA.1* \& BA.2, 10448–11287)}
\end{figure}

In Figure~\ref{fig:multiple_origins_graph} we give the details of these multiple pango lineages, along with an example lineage subgraph. The  earliest descendant sample of each of these putative independent XD origination nodes has EPI\_ISL: 11222324 (dated 2022-02-26) for the left hand plot, and EPI\_ISL 8514045 (dated 2021-12-30) for the right hand one. These two samples differ by  Supplementary Figure~\ref{fig:pango_XD_gisaid_graph} shows the equivalent subgraph with sample identifiers and GISAID Pango labels, none of which are XD (there are no samples designated by GISAID as XD in the long ARG).

@@

As well as cases where a single recombination event explains several X lineages (such as XAG above), many of the Pango X lineages involve more complex origins, in which samples which have not been previously identified as recombinants also descend from originating recombination events. In addition these complex topologies often include extra recombination nodes within the topology that descends from the main originating event. An example is XB in the long ARG, plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination between a B.1 sample and B.1.627 sample leads both to the XB samples and to a "hairball" of further recombination nodes whose descendants have not been identified as recombinants by Nextclade (blue nodes, "pre-Alpha" VoC). However, in this subgraph, there is only a single sample, identified as XB by Nextclade, which is descended from these
nested recombination events; moreover, this sample is not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph}). We suggest that investigating cases of complex origins, and identifing which (if any) of the nested, further recombination events may be artifactual, is an important area of future research.
